\documentclass[UTF8, a4paper]{article}

\usepackage{geometry}
\geometry{a4paper, margin=2.5cm}

\usepackage{amsmath, amssymb, amsthm}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{listings}
\lstset{
    basicstyle=\ttfamily\small,
    keywordstyle=\color{blue}\bfseries,
    commentstyle=\color{green!50!black}\itshape,
    stringstyle=\rmfamily\upshape,
    numbers=left,
    numberstyle=\tiny\color{gray},
    stepnumber=1,
    numbersep=5pt,
    backgroundcolor=\color{white},
    showstringspaces=false,
    breaklines=true,
    frame=single,
    rulecolor=\color{black},
}

\theoremstyle{definition}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}

\title{\bf{Design and Analysis of Directed Minimum Spanning Tree Algorithm}}
\author{Yanqiao Chen\\
        Southern University of Science and Technology\\
        Department of Computer Science and Engineering\\
        \texttt{12412115@mail.sustech.edu.cn}}
\date{\today}

\begin{document}

\maketitle

% ============================================================
% Section 1: Algorithm Description
% ============================================================
\section{Algorithm Description}

\subsection{Algorithm Overview}

This algorithm solves the Directed Minimum Spanning Tree (DMST) problem, also known as the minimum arborescence problem. We build upon the Chu-Liu/Edmonds algorithm and optimize it using a \textbf{mergeable heap} (leftist tree), \textbf{disjoint set union} (DSU), and \textbf{lazy propagation} to achieve $O(n \log m)$ time complexity.

What we do is to do a forward traversal from the root node (where we add dummy edges to ensure strong connectivity), and maintain 
a stack to keep track of the current path. For each node, we go to the source of its minimum incoming
edge and check: 
\begin{itemize}
    \item If it is in the same component as the current node, we pop it and continue. 
    \item If it is in a different component and not visited, we push it into the stack. 
    \item If it is in a different component and already visited, we have detected a cycle. We do the following: 
          \begin{itemize}
            \item Attribute a new index for the new super-node. 
            \item Backtrack the cycle nodes until we reach the source node again. 
            \item During the process, let the node be not visited, union the component to the new super-node 
            \item Then we update the lazy tags for the minimum incoming edge. 
            \item Add the weight to the answer 
            \item Pop out the selected edge, and the delete function will pushdown the lazy tags.
            \item Merge the heap of the cycle nodes into the new super-node.
          \end{itemize}
    \item After the cycle contraction, we push the new super-node into the stack for further exploration.
\end{itemize}

This will ends until the stack is empty, which means that all the components have been contracted into the root node. 
We will find the answer be greater or equal to $\infty$ if and only if the dummy edges are selected, which is 
a trick in the implementation to avoid boundary checks for nodes without valid incoming edges.

\subsection{Data Structures}

\subsubsection{Array-based Leftist Heap (Mergeable Heap)}

The leftist heap is a variant of the binary heap that supports efficient merging. To avoid the hidden overhead and memory fragmentation of dynamically allocating pointers via \texttt{new}, we employ an \textbf{array-based memory pool} for all heap nodes.

\paragraph{Heap Node Structure.} Each heap node stores:
\begin{itemize}
    \item \texttt{val}: original edge weight.
    \item \texttt{tag}: lazy propagation tag for subtracting edge weights when a cycle is formed.
    \item \texttt{from}, \texttt{to}: the endpoints of the directed edge.
    \item \texttt{lson}, \texttt{rson}: indices of the left and right children in the memory pool.
    \item \texttt{dis}: the distance to a null child, maintaining the leftist property.
\end{itemize}

\paragraph{Pushdown (Lazy Propagation).} Since edge weights of specific components need to be dynamically reduced when cycles are formed, a lazy tag (\texttt{tag}) is used. The \texttt{pushdown} operation adds the parent's \texttt{tag} to both left and right children's \texttt{val} and accumulates it into their \texttt{tag}s, then resets the parent's \texttt{tag} to zero.

\paragraph{Merge Operation.} The merge of two leftist heaps $x$ and $y$ is performed recursively:
\begin{enumerate}
    \item If either $x$ or $y$ is empty, return the non-empty one.
    \item Apply \texttt{pushdown} to both $x$ and $y$ to ensure values are up-to-date.
    \item Ensure the min-heap property by swapping $x$ and $y$ if $val(x) > val(y)$.
    \item Recursively merge $y$ with the right child of $x$: $rson(x) \gets merge(rson(x), y)$.
    \item Maintain the leftist property: if $dis(lson(x)) < dis(rson(x))$, swap $lson(x)$ and $rson(x)$.
    \item Update the distance: $dis(x) \gets dis(rson(x)) + 1$.
    \item Return $x$.
\end{enumerate}
This operation bounds the right spine length to $O(\log m)$, ensuring $O(\log m)$ merge time.

\paragraph{Delete (Pop) Operation.} To extract the minimum element (the root) and remove it, we simply call \texttt{pushdown} on the root, and then return the result of \texttt{merge(lson(root), rson(root))}.

\subsubsection{Other Data Structures}

\begin{itemize}
    \item \textbf{Disjoint Set Union} (\texttt{fa}): Tracks contracted super-nodes. Instead of overwriting existing vertices, newly contracted cycles are allocated strictly new node identifiers ($n+1, n+2, \dots$), ensuring a clean taxonomy.
    \item \textbf{Exploration Stack} (\texttt{sta}) and \textbf{Visited Array} (\texttt{vis}): Used to maintain the current path of forward traversal. If we query the minimum incoming edge and its source is already marked as visited in the current path, we have explicitly detected a cycle.
\end{itemize}

\subsection{Algorithm Workflow}

\begin{algorithm}[H]
\caption{Tarjan's $O(E \log V)$ DMST Algorithm with Stack and Dummy Edges}
\label{alg:dmst}
\begin{algorithmic}[1]
\Require Directed graph $G=(V, E)$, root node $r$
\Ensure Weight of minimum arborescence

\State Initialize array-based leftist heaps $root[v]$ for each $v \in V$
\For{each edge $(u, v, w) \in E$}
    \State Insert edge into $root[v]$
\EndFor
\For{$i = 1$ \textbf{to} $n$}
    \State Insert a dummy edge $(i, (i \bmod n) + 1, \infty)$ into $root[(i \bmod n) + 1]$
\EndFor
\State Initialize DSU $fa[i] \gets i$ for $i \in [1, 2n]$
\State $stack.push(r)$, $vis[r] \gets \text{true}$
\State $ans \gets 0$, $cnt \gets n$

\While{$root[stack.top()] \neq \emptyset$}
    \State $e \gets \text{minimum edge from } root[stack.top()]$
    \State $u \gets find(e.from)$
    \If{$u == stack.top()$}
        \State $root[stack.top()].pop()$ \Comment{Skip self-loops}
        \State \textbf{continue}
    \EndIf
    
    \If{\textbf{not} $vis[u]$}
        \State $stack.push(u)$, $vis[u] \gets \text{true}$ \Comment{Extend path}
        \State \textbf{continue}
    \EndIf
    
    \State $p \gets ++cnt$ \Comment{Cycle detected, create super-node}
    \While{$vis[u]$}
        \State $v \gets stack.pop()$
        \State $vis[v] \gets \text{false}$, $fa[v] \gets p$
        \State $val \gets root[v].top().val$
        \State $root[v].tag \gets root[v].tag - val$ \Comment{Lazy subtraction}
        \If{$find(root[v].top().to) \neq find(r)$}
            \State $ans \gets ans + val$ \Comment{Add edge weight to answer}
        \EndIf
        \State $root[v].pop()$ \Comment{Discard the selected cycle edge}
        \State $root[p] \gets merge(root[p], root[v])$
    \EndWhile
    \State $stack.push(p)$, $vis[p] \gets \text{true}$
\EndWhile

\State \Return $ans \geq \infty \;\mathbf{?}\; -1 : ans$
\end{algorithmic}
\end{algorithm}

\subsection{Key Design Decisions}

\paragraph{Dummy Edges for Strong Connectivity.} 
We proactively add $n$ edges with weight $\infty$ forming a simple cycle covering all nodes. This guarantees the graph is strongly connected, completely eliminating the need for boundary checks when a node has no valid incoming edges. If the final answer is greater than or equal to $\infty$, it implies the dummy edges were strictly necessary, hence no valid DMST exists in the original graph.

\paragraph{Stack-based Forward Traversal.}
Instead of maintaining predecessor arrays and iterating node-by-node, we use a stack to maintain the ``current exploration path''. We continuously inspect the incoming edges of the stack's top node. If an edge originates from a node already in the stack, a directed cycle is straightforwardly detected, and we pop the cycle nodes into a new super-node.

\paragraph{Array-Based Nodes and Clean IDs.}
When a cycle is contracted, rather than arbitrarily picking a representative among the cycle's nodes, we allocate a strictly new component ID ($n+1, n+2, \dots$ up to $2n-1$). This simplifies DSU structures and perfectly preserves the states of internal merged vertices.

% ============================================================
% Section 2: Complexity Analysis
% ============================================================
\section{Complexity Analysis}

\subsection{Heap Initialization}

We insert each of the $m$ edges into the heap of its destination node. For a leftist heap, insertion is equivalent to a merge operation with a single-node heap, which takes $O(\log m)$ time per edge. Therefore, the total initialization time is:
\[
T_{\text{init}} = O(m \log m)
\]

\subsection{Main Loop Analysis}

The main logic operates over a stack of vertices traversing the graph dynamically.

\paragraph{DSU operations.} Each $find$ operation takes $O(\alpha(n))$ time, effectively checking connected component identity efficiently.

\paragraph{Heap operations.} Extracting the minimum edge operates in $O(1)$, whilst deleting minimum elements and merging heaps take $O(\log m)$. Crucially, any valid directed edge from the graph is popped at most once. Merging heaps also occurs at most $n-1$ times because a new super-node strictly consumes $k$ old nodes and produces $1$ new node, decrementing total active connected components.
The total number of executed stack `pop' actions naturally scales linearly with the structural components count, thus summing heap merges limits them to $O(n \log m)$.

\paragraph{Cycle contraction bounds.} We start exactly with $n$ vertices. In each contraction, we discard internal minimal cycle edges. Because contractions reduce the count of strong components by strictly at least one, at most $n-1$ contractions will occur, thereby keeping internal allocations (e.g., node identifiers stepping toward $2n$) constrained within bounded linear scopes.

\subsection{Overall Time Complexity}

Combining all components:
\begin{align*}
T(n, m) &= T_{\text{init}} + T_{\text{traversal}} + T_{\text{contraction}} \\
&= O(m \log m) + O(n \cdot \alpha(n)) + O(n \log m) \\
&= O(m \log m) + O(n \log m)
\end{align*}

In a connected graph, $O(m \log m) = O(m \log n)$. Therefore, the overall time complexity is:
\[
\boxed{O(m \log n)}
\]

\subsection{Space Complexity}

The space complexity is dominated by the statically allocated arrays used to maintain the graph and the components:
\begin{itemize}
    \item \textbf{Leftist Heap Pool:} The array-based memory pool for the leftist tree stores exactly $m$ original edges plus $n$ dummy edges. Each node contains a constant number of integer fields (\texttt{lson}, \texttt{rson}, \texttt{val}, \texttt{tag}, \texttt{from}, \texttt{to}, \texttt{dis}), taking exactly $O(n + m)$ space.
    \item \textbf{Disjoint Set Union (DSU):} The \texttt{fa} array is initialized with a size of $2n$ to accommodate up to $n-1$ strictly new super-nodes generated during cycle contractions, consuming $O(n)$ space.
    \item \textbf{Auxiliary Arrays:} The \texttt{root}, \texttt{sta} (stack), and \texttt{vis} arrays similarly only need to track up to $2n$ total components throughout the algorithm's lifecycle, taking $O(n)$ space.
\end{itemize}

Combining these precisely bounded contiguous arrays, without any dynamic pointer allocation overhead, the overall space complexity is strictly:
\[
\boxed{O(n + m)}
\]

\appendix 

\section{Sample Code Implementation}

\begin{lstlisting}[language=C++, caption=Directed Minimum Spanning Tree Algorithm Implementation]
#include<iostream>
#include<cstring>
using namespace std;

#define int long long
const int N=1e6+10, INF=0x3f3f3f3f;

struct ltt_node {
    int lson, rson;
    int val, tag;
    int from, to;
    int dis;
};

struct leftist_tree {
    ltt_node ltt[N];
    int tot;
    
    inline int newnode(int val, int from, int to) {
        tot++;
        ltt[tot].val = val;
        ltt[tot].from = from;
        ltt[tot].to = to;
        return tot;
    }
    
    inline void pushdown(int now) {
        int ls = ltt[now].lson, rs = ltt[now].rson;
        if(ls) {
            ltt[ls].val += ltt[now].tag;
            ltt[ls].tag += ltt[now].tag;
        }
        if(rs) {
            ltt[rs].val += ltt[now].tag;
            ltt[rs].tag += ltt[now].tag;
        }
        ltt[now].tag = 0;
    }
    
    int merge(int x, int y) {
        if(!x || !y) return x + y;
        pushdown(x); pushdown(y);
        if(ltt[x].val > ltt[y].val) swap(x, y);
        ltt[x].rson = merge(ltt[x].rson, y);
        if(ltt[ltt[x].rson].dis > ltt[ltt[x].lson].dis)
            swap(ltt[x].lson, ltt[x].rson);
        ltt[x].dis = ltt[ltt[x].rson].dis + 1;
        return x;
    }
    
    int del(int rt) {
        pushdown(rt);
        int ls = ltt[rt].lson;
        int rs = ltt[rt].rson;
        return merge(ls, rs);
    }
}; 

leftist_tree ltt;
int root[N], fa[N];
int sta[N], top;
bool vis[N];

inline int find(int x) {
    return x == fa[x] ? x : fa[x] = find(fa[x]);
} 

signed main() {
    ios::sync_with_stdio(false);
    cin.tie(nullptr);
    
    int n, m, r;
    cin >> n >> m >> r;
    for(int i = 1; i <= m; i++) {
        int x, y, z;
        cin >> x >> y >> z;
        int lp = ltt.newnode(z, x, y); 
        root[y] = ltt.merge(root[y], lp); 
    }
    
    for(int i = 1; i <= n; i++) {
        int x = i, y = i % n + 1;
        int p = ltt.newnode(INF, x, y);
        root[y] = ltt.merge(root[y], p);
    } 
    
    for(int i = 1; i <= 2 * n; i++) fa[i] = i; 
    sta[++top] = r; 
    vis[r] = true;
    int ans = 0, cnt = n;
    
    while(root[sta[top]]) { 
        // Depth First Search Stack
        int lp = root[sta[top]]; // find the minimum edge in the heap
        ltt_node tmp = ltt.ltt[lp]; // get the edge info.
        int u = find(tmp.from); // find which component it belongs to
        
        if(u == sta[top]) {
            root[sta[top]] = ltt.del(root[sta[top]]);
            continue;
        } // if they are in the same component, invalid
        
        if(!vis[u]) {
            sta[++top] = u;
            vis[u] = true;
            continue;
        } // if they are in different components while u is not visited, we add u to the path
        
        // if they are in different components while u is already visited, we have detected a cycle, and we need to contract the cycle into a new super-node.
        int p = ++cnt; // attribute a new index for the super-node
        while(vis[u]) { 
            int v = sta[top--]; // backtrack the cycle nodes until we reach u again
            vis[v] = false; // reset the visited status for the next iteration
            fa[v] = p; // union the cycle nodes into the new super-node
            
            int val = ltt.ltt[root[v]].val; // the weight of the minimum incoming edge for node v
            ltt.ltt[root[v]].tag -= val; // updating lazy tags
            int x = find(ltt.ltt[root[v]].to); // find the component of the source node of the minimum incoming edge for node v
            ans += (x != find(r)) * val; // the edge do not goes to the root
            
            root[v] = ltt.del(root[v]); // delete while pushdown the root of v, which is the minimum incoming edge for v, since it has been selected as part of the cycle and will not be used anymore
            root[p] = ltt.merge(root[p], root[v]); 
        }
        
        sta[++top] = p; // push the new super-node into the stack for further exploration
        vis[p] = true; 
    }
    
    cout << (ans >= INF ? -1 : ans) << '\n';
    return 0;
}

\end{lstlisting}


\end{document}
