\documentclass[a4paper, UTF-8]{article}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{tikz}
\usetikzlibrary{automata, positioning, arrows}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors

\theoremstyle{plain}
\newtheorem{theorem}{\indent Theorem}[section]
\newtheorem{lemma}[theorem]{\indent Lemma}
\newtheorem{assumption}{\indent Assumption}[section]
\newtheorem{note}[theorem]{\indent Notation}
\newtheorem{proposition}[theorem]{\indent Proposition}
\newtheorem{corollary}[theorem]{\indent Corollary}
\newtheorem{definition}{\indent Definition}[section]
\newtheorem{example}{\indent Example}[section]
\newtheorem{remark}{\indent Remark}[section]
\newenvironment{solution}{\begin{proof}[\indent\bf Solution]}{\end{proof}}
\renewcommand{\proofname}{\indent\bf Proof}

\begin{document}

\title{\bf{Lab2 Answers}}

\author{Yanqiao Chen*\\ 
  Department of Computer Science and Engineering\\
  Southern University of Science and Technology\\
  \texttt{12412115@mail.sustech.edu.cn}
  }

\maketitle

\section{The Pseudocode for Lab2A}

The pseudocode of the Lab2A is shown as Algorithm \ref{alg:GSSC}.
\begin{algorithm}
\caption{Gale-Shapley Algorithm for Student-College Matching}
\label{alg:GSSC}
\begin{algorithmic}[1]
\Require Number of students $N$, number of colleges $M$, student preferences $s$, college capacities $c$
\Ensure Stable matching between students and colleges
\State Initialize $nxt[0 \dots N-1] \gets 0$ \Comment{Next college to apply for each student}
\State Initialize $college\_heaps[0 \dots M-1]$ as min-heaps \Comment{Store admitted (score, student) pairs}
\State Initialize queue $free\_students \gets \{0, 1, \dots, N-1\}$
\While{$free\_students$ is not empty}
    \State $stu \gets free\_students.pop()$
    \If{$nxt[stu] \geq |s[stu].pref|$}
        \State \textbf{continue} \Comment{No more colleges to apply}
    \EndIf
    \State $col \gets s[stu].pref[nxt[stu]] - 1$
 
    \State $nxt[stu] \gets nxt[stu] + 1$
    \State $score \gets c[col].comment[stu]$
    \If{$college\_heaps[col].size < c[col].cap$}
        \If{$score > 0$}
            \State $college\_heaps[col].push((score, stu))$
        \Else
            \State $free\_students.push(stu)$
        \EndIf
    \ElsIf{$college\_heaps[col]$ is not empty}
        \State $(worst\_score, worst\_stu) \gets college\_heaps[col].top()$
       
        \If{$score > worst\_score$}
            \State $college\_heaps[col].pop()$
            \State $college\_heaps[col].push((score, stu))$
            \State $free\_students.push(worst\_stu)$
        \Else
            \State $free\_students.push(stu)$
        \EndIf
    \Else
        \State $free\_students.push(stu)$
    \EndIf
\EndWhile
\For{$i = 0$ to $M-1$}
    \State Extract all students from $college\_heaps[i]$ and output
\EndFor
\end{algorithmic}
\end{algorithm}

\section{Problem 1: The Complexity Analysis}

Each student will apply to each college at most once in this algorithm. However, since we use min-heaps to maintain the admitted students for each college, each insertion or deletion takes $O(\log C)$ time, where $C$ is the maximum capacity of a college. Thus, the runtime in the worst case will be $O(MN \log C)$. Each student must at least apply once or choose not to apply (if the satisfaction score is all negative, e.g. in case 4 of the Online Judge, it can be implied that there exists a student whose scores are all negative), thus it will be $\Omega(N)$. Thus, the time complexity will be $O(MN \log C)$ and $\Omega(N)$.

For $N$ students, we use an array with length $M$ to store the satisfaction score, which is the dominant term. We also use a queue ($O(N)$) and $M$ priority queues (heaps) in the process ($O(MN)$). Thus the space complexity will be $O(MN)$.

\section{Problem 2: Stability of the Matching}

\begin{definition}[Stable Matching for Student-College Setting]
    A matching is \textbf{stable} if it satisfies the following conditions:
    \begin{enumerate}
        \item \textbf{Individual Rationality:} No student is matched to a college with non-positive satisfaction score, and no college admits a student with non-positive score.
        \item \textbf{No Blocking Pair:} There does not exist a pair $(s, c)$ such that:
        \begin{itemize}
            \item Student $s$ prefers college $c$ to his/her current matched college (or $s$ is unmatched but $c$ gives positive satisfaction to $s$);
            \item College $c$ prefers student $s$ to at least one of its currently admitted students (or $c$ has remaining capacity and $s$ gives positive satisfaction to $c$).
        \end{itemize}
    \end{enumerate}
\end{definition}

We prove that the extended GS algorithm outputs stable matching by contradiction. First, consider a student with negative satisfaction for all colleges. Such student will not enter the matching, thus no pair of instability is formed by this student. There are two unstable situations: 

\begin{itemize}
    \item The unstable pair is formed by two matched pairs.
    \item The unstable pair is formed by a matched pair and a standalone student.
\end{itemize}

\begin{remark}
    Two standalone students can be viewed as stable, since two null colleges give the same score to these students.
\end{remark}

\subsection{Case 1: Unstable Pair Between Two Matched Pairs} 

Assume there exist two matched pairs $(A, B)$ and $(C, D)$ that form an unstable situation, where $w(A, D) > w(A, B)$ and $w(D, A) > w(D, C)$. Here $C$ is the least preferred student currently admitted to college $D$ (if $C$ is not the least preferred, there exists a $C'$ that would also form an unstable pair). Since $w(A, D) > w(A, B)$, student $A$ applied to college $D$ before applying to college $B$ in the algorithm. We consider two cases: 
\begin{itemize} 
    \item \textbf{If $D$ is not full:} When $A$ applied to $D$, $A$ would have been admitted (since $w(D, A) > w(D, C) > 0$). Thus $A$ would not be released by $D$, contradicting that $A$ is matched to $B$.
    \item \textbf{If $D$ is full:} Since $w(D, A) > w(D, C)$, when $A$ applied to $D$, either (1) $D$ had capacity and admitted $A$, or (2) $D$ was full but preferred $A$ to its least preferred admitted student, so $A$ would be admitted and the least preferred student would be rejected. In either case, $C$ would not remain admitted to $D$ while $A$ is rejected, a contradiction.
\end{itemize}

\subsection{Case 2: Unstable Pair Between a Matched Pair and a Standalone Student} 

Assume there exists a matched pair $(A, B)$ and a standalone student $C$ that form an unstable situation, where $w(C, B) > w(C, \text{null}) = 0$ (i.e., $C$ prefers $B$ to being unmatched) and $w(B, C) > w(B, A)$ (i.e., $B$ prefers $C$ to $A$). Since $w(C, B) > 0$, student $C$ must have applied to college $B$ during the algorithm (students apply to all colleges with positive satisfaction in order of preference). We consider:

\begin{itemize}
    \item \textbf{If $B$ is not full when $C$ applied:} $C$ would have been admitted to $B$, contradicting that $C$ is standalone.
    \item \textbf{If $B$ is full when $C$ applied:} Since $w(B, C) > w(B, A)$ and $A$ is currently admitted to $B$, $C$ must have been rejected by $B$ in favor of students all preferred to $C$ (including $A$ or someone better). However, since $w(B, C) > w(B, A)$, if $C$ applied when $A$ was the worst admitted student, $C$ would have replaced $A$. This contradicts that $C$ remains standalone while $A$ stays matched to $B$.
\end{itemize}

\subsection{Conclusion}

According to the discussion, all the cases contradict with our assumption. Thus, the unstable pair cannot exists, the matching is a stable matching.

\section{Problem 3: Student's Optimality}

\begin{definition}[Student-Optimal Matching]
    A stable matching is \textbf{student-optimal} if every student is matched to the best college that he/she can possibly obtain in any stable matching. Formally, for each student $s$, let $\mu(s)$ be the college matched to $s$ in the student-optimal matching, and $\mu'(s)$ be the college matched to $s$ in any other stable matching. Then $w(s, \mu(s)) \geq w(s, \mu'(s))$ for all students $s$.
\end{definition}

We prove that the student-applying GS algorithm produces the student-optimal stable matching.

\begin{lemma}
    If a student $s$ is rejected by a college $c$ during the algorithm, then no stable matching can pair $s$ with $c$.
\end{lemma}

\begin{proof}
    We prove by strong induction on the sequence of rejections during the algorithm.
    
    \textbf{Base case:} Consider the first rejection in the algorithm. Suppose student $s$ is rejected by college $c$ at this step. Then $c$ must be full with students $d_1, d_2, \ldots, d_k$ (where $k$ is the capacity of $c$), and all these students are strictly preferred by $c$ to $s$, i.e., $w(c, d_i) > w(c, s)$ for all $i$.
    
    Suppose for contradiction that there exists a stable matching $\mu^*$ where $(s, c) \in \mu^*$. Since $c$ has capacity $k$ and prefers all $d_i$ to $s$, at least one $d_i$ must be matched to a college different from $c$ in $\mu^*$, say $(d_i, c') \in \mu^*$ with $c' \neq c$ (or $d_i$ is unmatched). In the algorithm, $d_i$ was admitted to $c$, so $w(d_i, c) > 0$. Also, $d_i$ applied to $c$ before being rejected, so $d_i$ prefers $c$ to any college applied to later (including $c'$ or remaining unmatched). Thus $w(d_i, c) > w(d_i, c')$ (or $w(d_i, c) > 0$ if unmatched).
    
    Now consider the pair $(d_i, c)$: $d_i$ prefers $c$ to its current match, and $c$ prefers $d_i$ to $s$. Thus $(d_i, c)$ is a blocking pair for $\mu^*$, contradicting stability.
    
    \textbf{Inductive step:} Assume the lemma holds for all rejections occurring before a certain rejection of student $s$ by college $c$. When $s$ is rejected by $c$, college $c$ is full with students $d_1, d_2, \ldots, d_k$ all strictly preferred to $s$, i.e., $w(c, d_i) > w(c, s)$ for all $i$.
    
    Suppose for contradiction that there exists a stable matching $\mu^*$ where $(s, c) \in \mu^*$. Then some $d_i$ is not matched to $c$ in $\mu^*$, say $(d_i, c') \in \mu^*$ with $c' \neq c$ or $d_i$ is unmatched. By the inductive hypothesis, $d_i$ has not been rejected by any college that is valid for $d_i$ before being admitted to $c$. Therefore, if $c'$ is valid for $d_i$ (i.e., can appear in a stable matching with $d_i$), then $d_i$ would not have been rejected by $c'$ before applying to $c$. This means $d_i$ prefers $c$ to $c'$ (or $c'$ is not valid for $d_i$, meaning no stable matching can pair them).
    
    In either case, $w(d_i, c) > w(d_i, c')$ (or $d_i$ unmatched and $w(d_i, c) > 0$). Combined with $w(c, d_i) > w(c, s)$, the pair $(d_i, c)$ forms a blocking pair for $\mu^*$, contradiction.
\end{proof}

\textbf{Main Proof:} We prove that the algorithm produces the student-optimal matching. Consider any student $s$ who is matched to college $\mu(s)$ in the algorithm output. If $s$ is unmatched, then $s$ was rejected by all colleges with positive satisfaction, and by the Lemma, no stable matching can pair $s$ with any acceptable college.

If $s$ is matched to $\mu(s) = c$, suppose for contradiction that there exists a stable matching $\mu'$ where $s$ is matched to $c'$ with $w(s, c') > w(s, c)$. Since $s$ prefers $c'$ to $c$, student $s$ applied to $c'$ before $c$ in the algorithm. For $s$ to end up at $c$, $s$ must have been rejected by $c'$ at some point. By the Lemma, no stable matching can pair $s$ with $c'$, contradicting that $\mu'$ is stable with $(s, c') \in \mu'$.

Therefore, no student can obtain a better college in any stable matching than what they obtain in the algorithm output. This proves that the student-applying GS algorithm produces the student-optimal stable matching. 

\end{document}